Qualitative Probabilistic Programming

### 2015-06-05 paper/paper.tex
changes changes

  We declared two types (\texttt{Point} and \texttt{Cluster}) and one
  random function (\texttt{getPoint}).  Type annotations are necessary for random
  functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
  points are 2-dimensional, and the type \texttt{Categorical(2)} expresses the
  fact that there are 2 possible clusters (so a cluster may be either 0 or 1).

@@

  The first cluster is at (6.3, 5.0) and the second is at (5.0, 1.6).  They both
  have a standard deviation of 0.54 in the x direction and 0.69 in the y direction.
  We could use these parameters to fill in the generative model:
  \begin{lstlisting}
var model = function() {
  var cluster = randomInteger(nclusters);
  return [gaussian(cluster == 0 ? 6.3 : 5.0, 0.54),
          gaussian(cluster == 0 ? 5.0 : 1.6, 0.69)];
};
  \end{lstlisting}
  This model is estimated to assign probability density $e^{-393}$ to the data, yielding a perplexity of $e^{-393/150} = 0.0728$.

@@

    To determine $\eta(x)$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  For the gaussian
    distribution, the sufficient statistics are $X$ and $X^2$ but the only feature is $X$.  For the
    categorical distribution \texttt{Categorical(n)}, the sufficient statistics
    and features are both $[X = 1], [X = 2], ..., [X=n - 1]$.
    The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.

@@

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
    This ensures that only the
    natural parameters that are features of $y$ are affected by $x$.

@@

    To evaluate performance, for each model, we:
    \begin{itemize}
      \item
        Randomly generate parameters $\theta$
      \item
        Generate datasets $x_{train}, x_{test}$ using $\theta$
      \item
        Estimate $\log P(x_{test} | \theta)$
      \item
        Use the EM algorithm to infer approximate parameters $\hat{\theta}$ from $x_{train}$
      \item
        Estimate $\log P(x_{test} | \hat{\theta})$
      \item
        Use the previous estimates to estimate the regret  $\log P(x_{test} | \hat{\theta}) - \log P(x_{test} | \theta)$
    \end{itemize}
    Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
    We use the Sequential Monte Carlo algorithm for this, as described in \cite{dippl}.

@@

  \subsection{Clustering}
{\small
\begin{lstlisting}
var Cluster = Categorical(3);
var Point = Vector(2, Double);
var getPoint = randFunction(Cluster, Point);

@@

The factor analysis model is very similar to the clustering model.  The main difference is that we replace the categorical \texttt{ClusterType} type with a vector type.  This results in the model attempting to find each point as an affine function of a vector of standard normal values.

@@

In this example, we use the unknown function \texttt{transFun} for state transitions and \texttt{obsFun} for observations.  This means that we will learn both the state transitions and the observation distribution.

@@

\subsection{Latent Dirichlet allocation}
{\small
\begin{lstlisting}
var maxWordsPerDocument = 100;
var Class = Categorical(3);
var Word = Categorical(10);
var classToWord = randFunction(Class, Word);

var model = function() {
  var whichClass = randomValue(Class);
  var nWords = observe('len' + docIndex, randomIntegerERP, maxWordsPerDocument);
  repeat(nWords, function(wordIndex) {
    observe(classToWord, whichClass);
  });
};
\end{lstlisting}
}

In this example, we use the unknown function \texttt{classToWord} to map classes to word distributions.  Note that each column of the matrix of parameters for \texttt{classToWord} will represent a categorical distribution over words, and there will be one column for each class.

@@

  We have found that it is possible to write many useful machine learning models as Quipp programs and then use generic algorithms for inference.  Furthermore, performance is <???>.  This should make it much easier for non-experts to write useful machine learning models.

  In the future, it will be useful to expand the set of types supported.  It is possible to define reasonable default distributions for non-recursive algebraic data types, and it may also be possible to define them for recursive algebraic data types using catamorphisms.  Also, it will be useful to create a more usable interface to infer parameters and perform additional data pracessing given these parameters.

>>>>

  We declared two types (\texttt{Point} and \texttt{Cluster}) and one
  random function (\texttt{getPoint}).  Type annotations are necessary for random
  functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
  points are vectors in $\mathbb{R}^2$, and the type \texttt{Categorical(2)} expresses the
  fact that there are 2 possible clusters (so a cluster may be either 0 or 1).

@@

  The first cluster is at (6.3, 5.0) and the second is at (5.0, 1.6).  They both
  have a standard deviation of 0.54 in the x direction and 0.69 in the y direction.
  We could use these parameters to fill in the generative model, giving us a webppl program:
  \begin{lstlisting}
var model = function() {
  var cluster = randomInteger(nclusters);
  return [gaussian(cluster == 0 ? 6.3 : 5.0, 0.54),
          gaussian(cluster == 0 ? 5.0 : 1.6, 0.69)];
};
  \end{lstlisting}
  This model is estimated to assign probability density $e^{-393}$ to the data, yielding a perplexity of $e^{-393/150} = 0.0728$.

@@

    To determine $\eta(x)$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  For the gaussian
    distribution, the sufficient statistics are $X$ and $X^2$ but the only feature is $X$.  For the
    categorical distribution \texttt{Categorical(n)}, the sufficient statistics
    and features are both $[X = 1], [X = 2], ..., [X=n - 1]$.
    The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.
    Features are selected so they correspond
    to natural parameters that can take on any real number.

@@

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
    to a sufficient statistic of $y$ that is not a feature and whose column is not 1.
    This ensures that only the
    natural parameters that are features of $y$ are affected by $x$.

@@

    To evaluate performance, for each model, we:
    \begin{itemize}
      \item
        Randomly generate parameters $\theta$
      \item
        Generate datasets $x_{train}, x_{test}$ using $\theta$
      \item
        Estimate $\log P(x_{test} | \theta)$
      \item
        Use the EM algorithm to infer approximate parameters $\hat{\theta}_i$ from $x_{train}$, for iteration $i = 1, 2, ...$
      \item
        Estimate $\log P(x_{test} | \hat{\theta}_i)$
      \item
        Use the previous estimates to estimate the regret $\log P(x_{test} | \hat{\theta}_i) - \log P(x_{test} | \theta)$
    \end{itemize}
    Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
    We use the Sequential Monte Carlo algorithm for this, as described in \cite{dippl}.

@@

  \subsection{Clustering}

{\small
\begin{lstlisting}
var Cluster = Categorical(3);
var Point = Vector(2, Double);
var getPoint = randFunction(Cluster, Point);

@@

The factor analysis model is very similar to the clustering model.  The main difference is that we replace the categorical \texttt{ClusterType} type with a vector type.  \texttt{randomValue(Factors)} will return a vector from the standard multivariate normal distribution.  This results in the model attempting to predict each point as an affine function of a vector of standard normal values, with Gaussian noise.

@@

In this example, we use the unknown function \texttt{transFun} for state transitions and \texttt{obsFun} for observations.  This means that we will learn both the state transition matrix and the observation matrix.

@@

\subsection{Latent Dirichlet allocation}
{\small
\begin{lstlisting}
var maxWordsPerDocument = 100;
var Topic = Categorical(3);
var Word = Categorical(10);
var topicToWord = randFunction(Topic, Word);

var model = function() {
  var whichClass = randomValue(Topic);
  var nWords = observe(randomIntegerERP, maxWordsPerDocument);
  repeat(nWords, function(wordIndex) {
    observe(classToWord, whichClass);
  });
};
\end{lstlisting}
}

For latent Dirichlet allocation, we use the unknown function \texttt{topicToWord} to map latent topics to word distributions.  We will learn a different categorical distribution for each topic.

@@

  It is possible to write many useful machine learning models as Quipp programs and then use a generic Monte Carlo expectation maximization algorithm for inference.  This algorithm infers parameters reasonably well.  Non-experts will find it easier to write useful machine learning models in this language compared to other probabilistic programming lanvguages.

  In the future, it will be useful to support additional types.  For example, we could add types for standard exponential families such as Poisson and Gamma.  Supporting disjoint union types is also straightforward.  Also, it will be useful to create a more usable interface to infer parameters and perform additional data processing given these parameters.
